---
title: "Reanalysis Johnson2007 - figures and code"
author: "Peter Hettegger"
date: "`r Sys.Date()`"
output:
  BiocStyle::pdf_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "left")
```

The `sessionInfo()` is provided at the end of the document.


```{r}
rm(list = ls())

library(GGally)
library(ggplot2)
library(limma)
library(sva)
library(qvalue)
library(randRotation)

```


File from https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures/tree/master/reanalysis/Johnson2007/data
See also Nygaard, V., Rodland, E. A. & Hovig, E. Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses. Biostatistics kxv027 (2015). doi:10.1093/biostatistics/kxv027


```{r}
edata <- read.table(file = "dataExample2.txt", header = TRUE,
                    sep = "\t", dec = ".", stringsAsFactors = FALSE)
```


File from https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures/tree/master/reanalysis/Johnson2007/data
See also Nygaard, V., Rodland, E. A. & Hovig, E. Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses. Biostatistics kxv027 (2015). doi:10.1093/biostatistics/kxv027

```{r}
pdata <- read.table(file = "sampleInfoExample2.txt", header = TRUE,
                    sep = "\t", dec = ".", stringsAsFactors = TRUE,
                    row.names = 1)

samps <- which(pdata$Type != "WT")

pdata <- pdata[samps,]
edata <- edata[,samps]
pdata$Type <- droplevels(pdata$Type)
pdata$Batch <- as.factor(pdata$Batch)

all.equal(rownames(pdata), colnames(edata))



# flooring to 1
edata[edata<1] <- 1
# take out data with too many low/missing values.
negativeprobesfilter <- rowSums(edata>1) >= (0.9*ncol(edata))
edata <- edata[negativeprobesfilter,]
# quantilenormalize
edata.quan <- normalizeBetweenArrays(log2(edata), method="quantile")
```


```{r}
# for debugging
debug = FALSE
if(debug) edata <- edata[1:1000,]
```


```{r}
#### ComBat - "p ComBat" values

mod.com <- model.matrix(~Type, pdata)
edata.com <- ComBat(edata.quan,
                    batch = pdata$Batch,
                    mod = mod.com)


mod.fit <- model.matrix(~Type, pdata)
fit1 <- lmFit(edata.com, design = mod.fit)
fit1 <- eBayes(fit1)

ps.com  <- topTable(fit1, number = Inf, sort.by = "none")$P.Value
fdr.com <- topTable(fit1, number = Inf, sort.by = "none")$adj.P.Val

sum(fdr.com<0.05)
sum(qvalue(ps.com)$qvalues < 0.05)
```


```{r}
#### limma batch as covariate - "p Limma (+batch)" values

mod.fit <- model.matrix(~Type + Batch, pdata)
fit2 <- lmFit(edata.quan, design = mod.fit)
fit2 <- eBayes(fit2)

ps.lim  <- topTable(fit2, coef = 2, number = Inf, sort.by = "none")$P.Value
fdr.lim <- topTable(fit2, coef = 2, number = Inf, sort.by = "none")$adj.P.Val

sum(fdr.lim<0.05)
sum(qvalue(ps.lim)$qvalues < 0.05)
```


```{r}
#### ComBat with random rotations - "p ComBat - 2000 rot." values

mod.fit <- model.matrix(~Type, pdata)

rr1 <- initBatchRandrot(edata.quan, mod.fit, 2, pdata$Batch)

statistic <- function(Y, batch, mod){
    edata.com <- sva::ComBat(Y,
                            batch = batch,
                            mod = mod, mean.only = FALSE)
    
    fit1 <- limma::lmFit(edata.com, design = mod)
    fit1 <- limma::eBayes(fit1)

    abs(limma::topTable(fit1, number = Inf, sort.by = "none")$t)
}
```


```{r message=FALSE, results='hide', warning=FALSE}
rs1 <- rotateStat(rr1, R = 2000, statistic = statistic, pdata$Batch,
                  mod.fit, parallel = TRUE)
```


```{r}
ps.rot <- pFdr(rs1)
fdr.rot <- p.adjust(ps.rot, "BH")

sum(fdr.rot<0.05)
sum(qvalue(ps.rot)$qvalues < 0.05)

ps <- cbind(ps.com, ps.lim, ps.rot = ps.rot[,1])
colnames(ps) <- c("p ComBat", "p Limma (+batch)", "p ComBat - 2000 rot.")

apply(ps, 2, function(i)sum(qvalue::qvalue(i)$qvalues < 0.05))
apply(ps, 2, function(i)sum(p.adjust(i, "BH") < 0.05))
```


```{r fig.width = 6, fig.height = 6}
#### p-vals scatterplot

df1 <- data.frame(ps)
colnames(df1) <- colnames(ps)

F1 <- function(...){
    ggally_points(..., size = 0.7)+
        scale_y_log10()+
        scale_x_log10()+
        geom_abline(slope = 1, intercept = 0, lty = 2, lwd = 0.5, col = "red")
}
lower.pan <- list(continuous = F1, combo = "facethist", discrete = "facetbar",
                  na = "na")

my.cor <- function(...)
    ggally_statistic(
      text_fn = 
        function(x,y)formatC(cor(log(x),log(y)), digits = 3, format = "f"),
      title = "Corr",
      sep = ":\n",...)

upper.pan <- list(continuous = my.cor, combo = "box_no_facet",
                  discrete = "count", na = "na")

ggpairs(df1, lower = lower.pan, upper = upper.pan, diag = NULL)
```


```{r fig.width = 3, fig.height = 6}
#### Histograms

df2 <- reshape2::melt(ps)

ggplot(df2, aes(x=value))+
    geom_histogram(colour="black", fill="lightgreen", binwidth =0.1,boundary=0)+
    facet_grid(Var2 ~ ., switch = "y")+
    geom_abline(slope = 0, intercept = nrow(ps)/10, lty = 2, col = "blue")+
    theme(axis.title.y = element_text(vjust=+3.3))
```


```{r fig.width = 4, fig.height = 3}
## Histograms 2

ind <- 1:15
h.com <- hist(ps.com, breaks = 100, plot = FALSE)
h.lim <- hist(ps.lim, breaks = 100, plot = FALSE)
h.rot <- hist(ps.rot, breaks = 100, plot = FALSE)

maxcount <- max(h.com$counts, h.lim$counts, h.rot$counts)

lab <- factor(rep(colnames(ps), rep(length(ind),3)), levels = colnames(ps))

df1 <- data.frame(
  mids = c(h.com$mids[ind], h.lim$mids[ind], h.rot$mids[ind]),
  counts = c(h.com$counts[ind], h.lim$counts[ind], h.rot$counts[ind]),
  lab = lab)

ggplot(df1, aes(x = mids, y = counts, colour = lab, lty = lab))+
    geom_line(lwd = 0.6) +
    xlab("p-value")+ ylab("Frequency")+
    theme(axis.title.y = element_text(vjust=+3.3)) +
    theme(legend.justification=c(1,1), legend.position=c(0.98, 0.98), 
          legend.title = element_blank())+
    theme(legend.key.size = unit(1.5,"line"))
```

# Session Info

```{r}
sessionInfo()
```
